Including EW and mixed QCD-EW corrections into PDF fits

### 2020-04-01 paper/sec-introduction.tex
Intro - revised

In the first respect, because Quantum Chromodynamics (QCD) dominates the 
interactions occurring within colliding protons at the LHC, much effort 
has been devoted to the computation of higher-order QCD corrections: 
fully-differential next-to-leading order (NLO) results, possibly matched to the
parton shower, are currently automated in various general-purpose
Monte Carlo generators~\cite{Gleisberg:2008ta,Alwall:2014hca,Bellm:2015jjp}
(see also ref.~\cite{Buckley:2011ms} for a review),
while an increasing number of next-to-next-to-leading order (NNLO) predictions 
are becoming available for processes with various degrees of inclusiveness
(see {\it e.g.}~\cite{Amoroso:2020lgh} and references therein). The computation
of higher-order corrections in the Electroweak (EW) theory has 
witnessed a comparatively less comprehensive progress. Nevertheless frameworks 
were developed in which the QCD and EW couplings are simultaneously treated as 
small parameters in the perturbative expansion, and the computation of 
theoretical predictions, accurate to NLO (including mixed-coupling QCD+EW 
terms), is automated~\cite{Biedermann:2017yoi,Frederix:2018nkq}.

In the second respect, contemporary PDF
sets~\cite{Harland-Lang:2014zoa,Ball:2017nwa,Hou:2019efy}
incorporate a significant amount of LHC data, which is analysed with NNLO QCD 
theory by default. No EW corrections are included in the theoretical 
description of the experimental observables to which PDFs are optimised,
except for QED effects if a photon PDF is 
determined~\cite{Schmidt:2015zda,Bertone:2017bme,Harland-Lang:2019pla}. 
The resulting relative PDF uncertainy --- 
which accounts only for the uncertainty of the data and of the 
methodology inherent to each PDF determination --- can be as low as 1\% at the 
EW scale~\cite{Ball:2017nwa}. Theoretical uncertainties, possibly of comparable 
size ({\it e.g.} from missing higher-order terms in the perturbative expansion),
can also be included in a PDF determination and represented into PDF 
uncertainties~\cite{AbdulKhalek:2019bux,AbdulKhalek:2019ihb}.
This leads to a noticeable improvement of PDF accuracy, in that the overall 
statistical quality of the fit increases, at the price of a modest 
deterioration of PDF precision, in that PDF uncertainties are slightly inflated.

The two respects are intertwined: PDFs provide a way of obtaining a prediction
for a given process in terms of other processes, therefore the accuracy and 
precision with which PDFs are able to make predictions crucially depend on 
the accuracy of the perturbative computation used in the PDF determination
(matched to the precision of the data). Electroweak (including mixed QCD+EW) 
effects should therefore naturally be included in a PDF determination. The 
reason is twofold. First, one expects NNLO QCD and NLO EW corrections to be of 
comparable size because, at the EW scale, the QCD and EW coupling constants 
become similar, $\alpha_s^2\sim \alpha$. If NNLO QCD corrections are included 
by default in PDF determinations, NLO EW and NLO EW+QCD corrections should be 
taken into account on the same footing. Second, the virtual exchange of soft or 
collinear weak bosons leads to Sudakov 
logarithms~\cite{Denner:2000jv,Denner:2001gw},
which can make the coefficients of the EW series grow faster than 
their QCD counterparts. Such a behaviour becomes relevant in
phase-space regions associated with large mass scales (roughly of the order
of the TeV), where several LHC data sets ({\it e.g.} $Z$ boson transverse 
momentum distributions) enter both the determination of PDFs and the search
for New Physics.

The systematic inclusion of EW corrections in a PDF determination entails
the solution of two separate problems. First, a problem of efficiency:
fast interpolation grids should be constructed, whereby partonic matrix 
elements, accurate to NLO QCD+EW, are precomputed in such a way that the 
numerical convolution with generic input PDFs can be efficiently approximated
by means of interpolation techniques. Such grids are a fundamental ingredient
in any PDF fit, where the evaluation of the hadronic cross section needs to 
be performed a large number of times. While two default formats exist for
these grids, {\sc APPLgrid}~\cite{Carli:2010rw} and 
{\sc fastNLO}~\cite{Kluge:2006xs}, none of them allows for the inlusion of EW
corrections nor for the interface to a Monte Carlo generator accurate to 
NLO QED+EW. Second, a problem of consistency: the way in which EW effects may
(or may not) be folded into the data varies across different experimental 
analyses, a fact that challenges their consistent theoretical interpretation. 
Examples are the subtraction of background processes which should not be 
considered as such ({\it e.g.} the $t$-channel photon-induced component in 
neutral-current Drell-Yan, which are not different processes beyond LO) or of 
just a part of the EW effects ({\it e.g.} multiple-photon radiation from light 
particles in the final state of neutral- or charged-current Drell-Yan, 
especially with electrons). Be that as it may, if EW effects are systematically
included in theoretical predictions, they should not be included in the data,
otherwise they will be double counted.

@@

The paper is organised as follows. In Sect.~\ref{sec:pineappl} we introduce
{\sc PineAPPL}, we describe its features, we illustrate its operation, and we
assess its performance. In Sect.~\ref{sec:results}, by means of 
{\sc mg5\_AMC@NLO} and {\sc PineAPPL}, we compute fast interpolation grids for 
a representative set of data that is typically included in a fit of PDFs.
In Sect.~\ref{sec:doublecounting} we use these results to formulate more 
comprehensively the second problem sketched above, the solution of which, 
however, remains beyond the scope of the current work. We provide an outlook in
Sect~\ref{sec:conclusion}.

>>>>

In the first respect, because Quantum Chromodynamics (QCD) dominates the 
interactions occurring within colliding protons at the LHC, much effort 
has been devoted to the computation of higher-order QCD corrections: 
fully-differential next-to-leading order (NLO) results, possibly matched to the
parton shower, are currently automated in various general-purpose
Monte Carlo generators~\cite{Gleisberg:2008ta,Alwall:2014hca,Bellm:2015jjp}
(see also~\cite{Buckley:2011ms} for a review),
while an increasing number of next-to-next-to-leading order (NNLO) predictions 
are becoming available for processes with various degrees of inclusiveness
(see {\it e.g.}~\cite{Amoroso:2020lgh} and references therein). The computation
of higher-order corrections in the Electroweak (EW) and mixed QCD+EW theory has 
witnessed a comparatively less comprehensive progress. Nevertheless frameworks 
were developed in which the QCD and EW couplings are simultaneously treated as 
small parameters in the perturbative expansion, and the computation of 
theoretical predictions, accurate to NLO in both (including mixed-coupling 
QCD+EW terms), is automated~\cite{Biedermann:2017yoi,Frederix:2018nkq}.

In the second respect, contemporary PDF
sets~\cite{Harland-Lang:2014zoa,Ball:2017nwa,Hou:2019efy}
incorporate a significant amount of LHC data, which is analysed with NNLO QCD 
theory by default. No EW corrections are systematically included in the 
theoretical description of the experimental observables to which PDFs are 
optimised, except for QED effects if a photon PDF is 
determined~\cite{Schmidt:2015zda,Bertone:2017bme,Harland-Lang:2019pla}. 
The resulting relative PDF uncertainy --- which accounts only for the 
uncertainty of the data and of residual methodology inefficiencies inherent to 
each PDF determination --- can be as low as 1\% at the EW 
scale~\cite{Ball:2017nwa}. Theoretical uncertainties, possibly of comparable 
size ({\it e.g.} from missing higher-order terms in the QCD perturbative 
expansion), have started to be included in a PDF determination and represented 
into PDF uncertainties only very 
recently~\cite{AbdulKhalek:2019bux,AbdulKhalek:2019ihb}. This has led to a 
noticeable improvement of PDF accuracy, in terms of an overall
better statistical quality of the fit, at the price of a modest deterioration 
of PDF precision, in terms of a slight inflation of PDF uncertainties.

The two respects are intertwined: PDFs provide a way of obtaining a prediction
for a given process in terms of other processes, therefore the accuracy and 
precision with which PDFs are able to make predictions crucially depend on 
the accuracy of the perturbative computation used in the PDF determination
(matched to the precision of the data). Electroweak (including mixed QCD+EW) 
effects should therefore naturally be included in a PDF determination. The 
reason is twofold. First, one expects NNLO QCD and NLO EW corrections to be of 
comparable size because, at the EW scale, the QCD and EW running couplings 
become similar, $\alpha_S^2\sim \alpha$. If NNLO QCD corrections are included 
by default in PDF determinations, NLO EW and NLO EW+QCD corrections should be 
taken into account on the same footing. Second, the virtual exchange of soft or 
collinear weak bosons leads to Sudakov 
logarithms~\cite{Denner:2000jv,Denner:2001gw},
which can make the coefficients of the EW series grow faster than 
their QCD counterparts. Such a behaviour becomes relevant in
phase-space regions associated with large mass scales (roughly of the order
of the TeV), where several LHC data sets ({\it e.g.} $Z$ boson transverse 
momentum distributions) enter both the determination of PDFs and the search
for New Physics.

The systematic inclusion of EW corrections in a PDF determination entails
the solution of two separate problems. First, a problem of efficiency:
fast interpolation grids should be constructed, whereby partonic matrix 
elements, accurate to NLO QCD+EW, are precomputed in such a way that the 
numerical convolution with generic input PDFs can be efficiently approximated
by means of interpolation techniques. Such grids are a fundamental ingredient
in any PDF fit, where the evaluation of the hadronic cross section needs to 
be performed a large number of times. While two default formats exist for
these grids, {\sc APPLgrid}~\cite{Carli:2010rw} and 
{\sc fastNLO}~\cite{Kluge:2006xs}, none of them supports the inclusion of EW
corrections nor the interface to a Monte Carlo generator accurate to 
NLO QED+EW. Second, a problem of consistency: the way in which EW effects may
(or may not) be folded into the data varies across different experimental 
analyses, a fact that challenges their consistent theoretical interpretation. 
Examples are the subtraction of background processes which should not be 
considered as such ({\it e.g.} the $t$-channel photon-induced component in 
neutral-current Drell-Yan, which is not a separate processes beyond the leading
order) or of just a part of the EW effects ({\it e.g.} multiple-photon 
radiation from light particles in the final state of neutral- or charged-current
Drell-Yan, especially with electrons). Be that as it may, if EW effects are 
systematically included in theoretical predictions, they should not be included 
in the data, otherwise they will be double counted.

@@

The paper is organised as follows. In Sect.~\ref{sec:pineappl} we introduce
{\sc PineAPPL}, we describe its features, we illustrate its operation, and we
assess its performance. In Sect.~\ref{sec:results}, by means of 
{\sc mg5\_AMC@NLO} and {\sc PineAPPL}, we compute fast interpolation grids for 
a representative set of data that is typically included in a fit of PDFs.
In Sect.~\ref{sec:doublecounting} we use these results to formulate more 
comprehensively the second problem sketched above, the solution of which, 
however, remains beyond the scope of the current work. We provide our
conclusions and an outlook in Sect.~\ref{sec:conclusion}.
